\chapter{Stochastic Resonance in a Spiking Neural Network Model of the Honeybee Antennal Lobe}
\section{Introduction}

One of the first well established findings in neuroscience with the advent of electrical recordings of neuronal activity, was the variability of neurons' responses. 
That is, when stimulated with the same input, neurons respond with different temporal patterns of spikes, while the overall firing rate remains roughly equal \cite{tomkoNeuronalVariabilityNonstationary1974,mainenReliabilitySpikeTiming1995,tolhurstStatisticalReliabilitySignals1983}.

This trial-to-trial variability is commonly called noise. 
Sources of noise in neurons are many and not all variability in response is due to noise (see \parencite{faisalNoiseNervousSystem2008} for a review). 
Still, due to the physical scale of cells, a stochastic component in neuronal function is inevitable by thermal noise alone. % cite something
It is then to be expected that neural systems would have evolved to tolerate or, more daringly, exploit the presence of noise. 

\subsection{Classical Stochastic Resonance}
The idea of studying the effects of noise on the dynamics of a system first appeared in the work of \textcite{benziMechanismStochasticResonance1981,benziStochasticResonanceClimatic1982}, where the term Stochastic Resonance (SR) was used to explain the periodic behavior of earth's average temperature. 
A brief covering of that first SR application will be made, as it provides a practical understanding of the general phenomenon and a surprising link to neural systems.

The most common way to model temperature averages was using simple energy balance deterministic models:

\begin{equation}
    \label{eq:temp model}
    C\odv{T}{t} = R_{in}(T) - R_{out}(T)
\end{equation}

where $C$ is the earth's thermal capacity, and $R_{in}$ and $R_{out}$ represent the total incoming and outgoing solar radiation, defined as

\begin{align}
    &R_{in} = Q\mu \\
    &R_{out} = \alpha(T)Q\mu + \varepsilon(T)
\end{align}

where $\mu$ is the average incoming solar radiation.

In \cite{benziStochasticResonanceClimatic1982} a simplification is used: 3 steady-states are assumed to exist as $T_1 < T_2 < T_3$, those are possible earth's climates.
\ref{eq:temp model} then becomes

\begin{equation}
    C\odv{T}{t} = \frac{\mu\varepsilon(T)}{1 + \beta \left( 1 - T/T_1 \right) \left( 1 - T/T_2 \right) \left( 1 - T/T_3 \right)} -\varepsilon (T)
\end{equation}
%should be explained better

To analyze the stability of the defined climates the rate of change of temperatures is taken as

\begin{equation}
    F(T) = \odv{T}{t} \text{ , }
\end{equation}
then a pseudo-potential energy function is defined as

\begin{equation}
    \Phi(T) = - \int F(T) \mathrm{d}T
\end{equation}

that represents the "energy" needed to change the temperature across temperature values. % this should be explained more clearly
From $\Phi$ we get

\begin{equation}
    F(T) = - \pdv{\Phi}{T}
\end{equation}

Then intuitively the maximum of $\Phi$ will be an unstable point, while the minima will be stable. That is, where the second derivative with respect to temperature

\begin{equation}
    \odv{F}{T} = - \pdv[2]{\Phi}{T}
\end{equation}

is $<0$, any neighboring starting state of the earth's temperature will converge to it, while when it is $>0$ they will diverge.

It can be shown that $T_1 \text{, } T_3$ are stable points, while $T_2$ is unstable: a peak any starting state would ''roll off'' from (see fig). % add an appendix showing that?
In our context, $T_1 \text{, } T_3$ are the only observable climates, temperatures at which the earth will stay for enough time to be detectable in the geological records.
This type of system where 2 stable states exist is called a bistable system, and the plot of $\Phi$ over $T$ displays the typical double well shape of the (pseudo)potential function. % add plot

With the present model, we can see that the earth's climate would depend solely (all else being equal) on the starting temperature, i.e., wether it is smaller or greater than $T_2$.
Then it would follow $F(T)$ towards $T_1$ or $T_3$. However, earth's temperatures do indeed show oscillations with a period of $10^5$ years.

A weak cosinusoidal forcing modulating the input solar radiations can be introduced, representing the varying distance of the earth from the sun along its elliptic orbit. 
Then $\mu$ becomes time dependent

\begin{equation}
    \mu(t) = 1 + \gamma \cos(\omega t)
\end{equation}
where $\gamma \ll 1$ and $\omega = 2\pi/10^5$, reflecting the observed oscillation period.
The derivative and pseudo-energy function become $\tilde{F} = \tilde{F}(T;t) \text{, } \tilde{\Phi} = \tilde{\Phi}(T;t)$.

Still, this periodicity alone does not lead to any change in climates: the system never jumps over $T_2$ (see fig). % add phase plane

Here the turning point, if gaussian white noise, representing various mechanisms causing $T$ fluctuations at a smaller time scale, is added to the system, yielding

\begin{equation}
    \odv{T}{t} = F(T) + \sigma \eta(t)
\end{equation}
then for certain values of $\sigma$ a jump between stable states becomes possible, with a frequency $\approx 2\pi/10^5$.
That is, temperature oscillations across the geological record 

% just placeholder to understand the trajectory to be followed
% Then, in a "sensing system" sense, the (relatively) weak $\xi$ helped to recover the weak forcing frequency $\omega$, which would have not been detected otherwise. 
% In this sense noise may be seen as beneficial in signal processing, neurons of the first part of the sensory systems are particularly noisy signal processors.

